Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-08-16 submission/manuscript.Rmd
merging paragraphs

### Combining sections

In the above study, we demonstrate KDM5C's pivotal role in the development of tissue identity, resulting in substantial misexpression of tissue-enriched genes within the adult _Kdm5c_-KO hippocampus and amygdala. In addition to testis genes identified previously[@scandagliaLossKdm5cCauses2017], we found significant enrichment of muscle, liver, and even ovary-biased genes aberrantly expressed within the male _Kdm5c_-KO brain. Given the majority of tissue-enriched DEGs have no known brain functions, it is currently unknown if they impair _Kdm5c_-KO neurodevelopment and contribute to MRXSCJ-related phenotypes like intellectual disability and aberrant aggression. However, select liver and muscle-biased DEGs do have known roles within the brain, such as the liver-enriched lipid metabolism gene _Apolipoprotein C-I (Apoc1)_ [@roulandRoleApolipoproteinC12022] that is highly expressed in the _Kdm5c_-KO amygdala and hippocampus. Intriguingly, _Apoc1_ overexpression in the mouse brain can impair learning and memory[@abildayevaHumanApolipoproteinCI2008] and is implicated in Alzheimer's disease in humans[@leducAPOECholesterolHomeostasis2010], however further investigation is required to determine its impact upon _Kdm5c_-KO phenotypes.

Analysis of epiblast-like cells (EpiLCs) revealed dysregulation of _Kdm5c_-KO tissue identity begins during early emrbyogenesis. _In vivo_, germline genes are typically fully decommissioned in epiblast stem cells and remain silenced as the epiblast differentiates into somatic tissues[@bardotMouseGastrulationCoordination2020]. However, a small subset of epiblast stem cells will reactivate germline genes to become the primoridal germ cells (PGCs) that will ultimately form the mature germline[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014]. This process can be mimicked _in vitro_ by differentiating EpiLCs into primordial germ cell-like cells (PGCLCs)[@hayashiReconstitutionMouseGerm2011]. Thus, misexpression of germline genes in _Kdm5c_-KO EpiLCs might suggest ESCs are progressing beyond EpiLC differentiation and instead becoming PGCLCs. Yet, _Kdm5c_-KO EpiLCs properly express genes for primed pluripotency, including _Otx2_ - an epiblast stem cell marker that is known to repress differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018]. Furthermore, we observed no difference in cellular morphology during _Kdm5c_-KO ESC to EpiLC differentiation. Proper EpiLC differentiation, coupled with _Kdm5c_-KO mice being viable, suggests germline gene expression is occuring ectopically in conjunction with typical developmental programs, rather than a complete shift to towards germline identity.

While many genes involved in early germline development and meiosis are common between males and females, some germline genes display sex-biased expression or perform functions unique to eggs and sperm. We found both males and females expressed both sperm and egg biased genes, indicating the mechanisms underlying germline gene repression are independent of germ cell sex. However, we found organismal sex did greatly influence the manifestation of germline gene dysregulation, as _Kdm5c_-KO females had over double the number of germline-enriched DEGs compared to male EpiLCs. Knockout of KDM5C in females is embryonic lethal, likely due to impaired X chromosome inactivation (XCI) [@samantaActivationXistEvolutionarily2022]. XCI defects could explain why _Kdm5c_-KO females are more prone to germline gene dysregulation, given the X chromosome harbors a large number of spermatogenesis genes[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004]. However, ectopic germline transcripts, including those unique to females, were not significantly biased towards the X chromsome. Sex differences in germline gene suppression may be instead connected to females having a higher dose of KDM5C than males, due to its partial escape from XCI[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008]. Intriguingly, females heterozygous for _Kdm5c_ loss also had over double the number of germline DEGs than males, even though their XCI is largely preserved and their expression of KDM5C should be roughly equivalent to that of wild-type males. Altogether, this suggests female EpiLCs are more prone to transitioning to a germ cell-like state than males and require more KDM5C to maintain somatic cellular identity. Future studies investigating sex differences in germline gene repression are required to illuminate if this phenomenon is unique to sexually dimorphic chromatin regulators like KDM5C or a general feature of female cells.

@@

It is important to note that while we focused on KDM5C's regulation of germline genes, some germline-enriched genes are also expressed in na&#239;ve ESCs and at the 2-cell stage due to their roles in pluripotency and self-renewal. For example, _Dazl_ is primarily known for committing PGCs to the germline fate and regulating the translation of germline-specific RNAs, but is also expressed at the 2-cell stage[@yangChemicalinducedChromatinRemodeling2022], in na&#239;ve ESCs[@wellingDAZLRegulatesTet12015], and in the inner cell mass [@wellingDAZLRegulatesTet12015]. Very recently, two screens of _Dazl_-repressors identified KDM5C as a direct regulator in ESCs[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023]. Interestingly, one study found _Kdm5c_-KO ESCs also aberrantly expressed 2-cell-specific genes, indicating KDM5C negatively regulates totipotency[@guptaGenomewideScreenReveals2023]. However, out of the four regulators characterized, KDM5C was the only factor whose repression of _Dazl_ was independent of the 2-cell specific transcription factor _Dux_[@guptaGenomewideScreenReveals2023]. Although expressed at the 2-cell stage and in ESCs, _Dazl_ is typically silenced when ESCs differentiate into EpiLCs[@mochizukiRepressionGermlineGenes2021]. We found KDM5C also repressess _Dazl_ in EpiLCs, but _Kdm5c_-KO EpiLCs did not express 2-cell specific genes like _Dux_ and _Zscan4c_. Together, this suggests the 2-cell-like state in _Kdm5c_-KO ESCs is an indirect consequence of KDM5C's primary role in germline gene repression.

@@

The above work provides the mechanistic foundation for KDM5C's role in germline gene repression and offers novel insight into how dysregulation in _Kdm5c_-KO tissue identity changes over the course of development. However, the contribution of ectopic, tissue-specific genes towards _Kdm5c_-KO neurological impairments is still unknown. KDM5C may be crucial for neurodevelopment by fine-tuning the expression of tissue-erniched, dosage-sensitive genes, such as _Apoc1_. KDM5C could also ensure tissue-specific timing of transcription factor family members, like RFX genes that are broadly required for cilia and flagella formation, including neuronal cilia[@swobodaRFXTypeTranscriptionFactor2000; @ashiqueRfx4TranscriptionFactor2009; @chungRFX2BroadlyRequired2012; @coschieraPrimaryCiliaPromote2024]. While their impact upon neurodevelopment is currently unclear, ectopic germline transcripts are also found in models of other related neurodevelopmental disorders[@bonefasSomatogermlineTransformationChromatinlinked2021], including Immunodeficiency, Centromere instability, Facial anomalies syndrome 1 (ICF1 OMIM: #242860)[@velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014], Kleefstra syndrome 1 (OMIM: #610253)[@schaeferControlCognitionAdaptive2009], and MeCP2 duplication syndrome (MDS, OMIM: #300260)[@samacoCrhOprm1Mediate2012]. Like KDM5C, the chromatin regulators underlying these conditions, DNA methyltransferase 3b (DNMT3B), H3K9me1/2 methyltransferases G9A/GLP, methyl-CpG -binding protein 2 (MECP2), primarily silence gene expression. Thus, KDM5C is among a growing cohort of chromatin-linked neurodevelopmental disorders that have a similar underlying phenotype of germline versus soma dysregulation. Further research is required to determine the ultimate impact of these germline genes and the extent to which this phenomenon occurs in humans.

>>>>

It is important to note that while we highlighted KDM5C's regulation of germline genes, some germline-enriched genes are also expressed in na&#239;ve ESCs and at the 2-cell stage due to their roles in pluripotency and self-renewal. For example, _Dazl_ is primarily known for committing PGCs to the germline fate and regulating the translation of germline-specific RNAs, but is also expressed at the 2-cell stage[@yangChemicalinducedChromatinRemodeling2022], in na&#239;ve ESCs[@wellingDAZLRegulatesTet12015], and in the inner cell mass [@wellingDAZLRegulatesTet12015]. Emerging evidence suggests KDM5C might negatively regulate totipotency genes, as KDM5C suppresses _Dazl_ in ESCs[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023] and _Kdm5c_-KO ESCs aberrantly express 2-cell-specific genes like _Zscan4c_[@guptaGenomewideScreenReveals2023]. However, out of the four _Dazl_ regulators characterized, KDM5C was the only factor whose repression of _Dazl_ was independent of the 2-cell specific transcription factor _Dux_[@guptaGenomewideScreenReveals2023]. We found that while KDM5C also repressess _Dazl_ in EpiLCs, _Kdm5c_-KO EpiLCs did not express 2-cell specific genes. This suggests the 2-cell-like state observed in _Kdm5c_-KO ESCs is an indirect consequence of KDM5C's primary role in germline gene repression.

Although expressed at the 2-cell stage an in ESCs, _Dazl_ and other "self-renewal" germline genes are typically silenced when ESCs differentiate into EpiLCs[@mochizukiRepressionGermlineGenes2021], but can become reactivated to form primordial germ cell-like cells (PGCLCs)[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014; @hayashiReconstitutionMouseGerm2011].

Yet, _Kdm5c_-KO EpiLCs properly express genes for primed pluripotency, including _Otx2_ - an epiblast stem cell marker that is known to repress differentiation into PGCs/PGCLCs[@zhangOTX2RestrictsEntry2018]. Proper _Otx2_ expression, coupled normal EpiLC cellular morphology and _Kdm5c_-KO mice being viable, suggests germline gene expression is occuring ectopically in conjunction with typical developmental programs, rather than a complete shift to towards germline identity.

misexpression of germline genes in  _Kdm5c_-KO EpiLCs might suggest mutant ESCs are instead differentiating into primordial germ cell-like cells

Since epiblast stem cells can reactivate germline genes and differentiate into primoridal germ cells[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014], misexpression of germline genes in  _Kdm5c_-KO EpiLCs might suggest mutant ESCs are instead differentiating into primordial germ cell-like cells (PGCLCs)[@hayashiReconstitutionMouseGerm2011].

### males vs females

While many genes involved in early germline development and meiosis are common between males and females, some germline genes display sex-biased expression or perform functions unique to eggs and sperm. We found both males and females expressed both sperm and egg biased genes, indicating the mechanisms underlying germline gene repression are independent of germ cell sex. However, we found organismal sex did greatly influence the manifestation of germline gene dysregulation, as _Kdm5c_-KO females had over double the number of germline-enriched DEGs compared to male EpiLCs. Knockout of KDM5C in females is embryonic lethal and results in impaired X chromosome inactivation (XCI) [@samantaActivationXistEvolutionarily2022]. XCI defects could explain why homozygous _Kdm5c_ knockout females are more prone to germline gene dysregulation, given the X chromosome harbors a large number of spermatogenesis genes[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004]. However, ectopic germline transcripts, including those unique to females, were not significantly biased towards the X chromsome. Sex differences in germline gene suppression may be instead connected to females having a higher dose of KDM5C than males, due to its partial escape from XCI[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008]. Intriguingly, heterozygous loss of _Kdm5c_ in females also had over double the number of germline DEGs than males, even though their expression of KDM5C should be roughly equivalent to that of wild-type males. Altogether, these results suggest female EpiLCs are more prone to transitioning to a germ cell-like state than males and require more KDM5C to maintain somatic cellular identity. Future studies investigating sex differences in germline gene repression are required to illuminate if this phenomenon is unique to sexually dimorphic chromatin regulators or a general feature of female cells.

### ChIPseq motifs and transcription factors

@@

### Germline gene decommissioning dna methylation
- End with comment on catalytic dependent vs independent functions of chromatin regulators
	- Requirement for catalytic activity may change depending upon the context

@@

### Functional consequences 
In addition to testis genes identified previously[@scandagliaLossKdm5cCauses2017], we found significant enrichment of muscle, liver, and even ovary-biased genes aberrantly expressed within the male _Kdm5c_-KO brain. Given the majority of tissue-enriched DEGs have no known brain functions, it is currently unknown if they impair _Kdm5c_-KO neurodevelopment and contribute to MRXSCJ-related phenotypes like intellectual disability and aberrant aggression. However, select liver and muscle-biased DEGs do have known roles within the brain, such as the liver-enriched lipid metabolism gene _Apolipoprotein C-I (Apoc1)_ [@roulandRoleApolipoproteinC12022] that is highly expressed in the _Kdm5c_-KO amygdala and hippocampus. Intriguingly, _Apoc1_ overexpression in the mouse brain can impair learning and memory[@abildayevaHumanApolipoproteinCI2008] and is implicated in Alzheimer's disease in humans[@leducAPOECholesterolHomeostasis2010], however further investigation is required to determine its impact upon _Kdm5c_-KO phenotypes.

The above work provides the mechanistic foundation for KDM5C's role in germline gene repression and offers novel insight into how dysregulation in _Kdm5c_-KO tissue identity changes over the course of development. However, the contribution of ectopic, tissue-specific genes towards _Kdm5c_-KO neurological impairments is still unknown. KDM5C may be crucial for neurodevelopment by fine-tuning the expression of tissue-erniched, dosage-sensitive genes, such as _Apoc1_. KDM5C could also ensure tissue-specific timing of transcription factor family members, like RFX genes that are broadly required for cilia and flagella formation, including neuronal cilia[@swobodaRFXTypeTranscriptionFactor2000; @ashiqueRfx4TranscriptionFactor2009; @chungRFX2BroadlyRequired2012; @coschieraPrimaryCiliaPromote2024]. While their impact upon neurodevelopment is currently unclear, ectopic germline transcripts are also found in models of other related neurodevelopmental disorders[@bonefasSomatogermlineTransformationChromatinlinked2021], including Immunodeficiency, Centromere instability, Facial anomalies syndrome 1 (ICF1 OMIM: #242860)[@velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014], Kleefstra syndrome 1 (OMIM: #610253)[@schaeferControlCognitionAdaptive2009], and MeCP2 duplication syndrome (MDS, OMIM: #300260)[@samacoCrhOprm1Mediate2012]. Like KDM5C, the chromatin regulators underlying these conditions, DNA methyltransferase 3b (DNMT3B), H3K9me1/2 methyltransferases G9A/GLP, methyl-CpG -binding protein 2 (MECP2), primarily silence gene expression. Thus, KDM5C is among a growing cohort of chromatin-linked neurodevelopmental disorders that have a similar underlying phenotype of germline versus soma dysregulation. Ultimately, further research is required to determine the impact of these germline genes and the extent to which this phenomenon occurs in humans.

Analysis of epiblast-like cells (EpiLCs) revealed dysregulation of _Kdm5c_-KO tissue identity begins during early emrbyogenesis. _In vivo_, germline genes are typically fully decommissioned in epiblast stem cells and remain silenced as the epiblast differentiates into somatic tissues[@bardotMouseGastrulationCoordination2020]. However, a small subset of epiblast stem cells will reactivate germline genes to become the primoridal germ cells (PGCs) that will ultimately form the mature germline[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014]. This process can be mimicked _in vitro_ by differentiating EpiLCs into primordial germ cell-like cells (PGCLCs)[@hayashiReconstitutionMouseGerm2011].
